How fluid-mechanical erosion creates anisotropic porous media

### 2022-11-30 docs/PhysicsPaper/StokesErosion.tex
NM

There are two mechanistic explanations for how erosion can create medium anisotropy. First, the shear stresses could carve each individual body into a more slender form, thus creating anisotropy at the level of individual grains. Second, the shear stressed could preferentially remove certain bodies before others, thus creating large-scale anisotropy. For example, a body positioned in a high-throughput channel might disintegrate relatively quickly, thus opening the channel and creating higher overall anisotropy. We will refer to these two possible mechanisms as {\em shape} anisotropy and {\em configurational} anisotropy, respectively. As an extreme example, a tightly-packed horizontal row of circular bodies would exhibit high configurational anisotropy but no shape anisotropy. Meanwhile, an array of highly eccentric ellipses, all oriented horizontally but positioned randomly, would exhibit high shape anisotropy and low configurational anisotropy. We note that channelization is likely associated with {\em both} types of anisotropy. Certainly the overall configuration must support a channel, and, secondly, the geometry of the individual bodies that outline the channel could control its structure to some degree.

@@

As a second way to quantify medium anisotropy, we measure the
tortuosity, $T_1$ and $T_2$, in the longitudinal and transverse
directions respectively. Figure~\ref{fig4}(a), shows how both vary over
the course of the simulation. As the bodies erode, $T_2$ grows
significantly larger than $T_1$, indicating that passive tracers must
follow more tortuous paths when traveling transversely compared to longitudinally. The ratio $(T_2-1)/(T_1-1)$ provides
a second measure of anisotropy to complement the permeability-dependent
quantity. We have chosen to subtract one in parenthesis so that both
numerator and denominator both vanish when no bodies are present, as
analogous to resistivity. As seen in Figure~\ref{fig4}(b), the ratio $(T_2-1)/(T_1-1)$ grows significantly during erosion, up to a peak of nearly 8. This ratio can be decomposed into configurational and shape components, exactly as is done for the anisotropy (i.e.~by replacing a collection of eroded bodies with circles of the same areas and centers of mass). Figure~\ref{fig4}(b) shows that the shape component is greater for the majority of the simulation, but near the end of the simulation, the configurational component is competitive.

@@

The previous section shows intriguing features to arise from the erosion of a single, random initialization of solid bodies. The observations natural raise the question: are the trends specific to the particular simulation shown or do these features emerge robustly across different initial conditions? To answer this question, we now analyze the same quantities---permeability, tortuosity, and anisotropy---over an ensemble of simulations having different numbers of bodies and different random initializations. We analyze runs having $M$ = 20, 40, 60, 80, and 100 bodies, with at least three different instances of each. This data set represents roughly 2,100 hours of computational time.

>>>>

There are two mechanistic explanations for how erosion can create medium anisotropy. First, the shear stresses could carve each individual body into a more slender form, thus creating anisotropy at the level of individual grains. Second, the shear stresses could preferentially remove certain bodies before others, thus creating large-scale anisotropy. For example, a body positioned in a high-throughput channel might disintegrate relatively quickly, thus opening the channel and creating higher overall anisotropy. We will refer to these two possible mechanisms as {\em shape} anisotropy and {\em configurational} anisotropy, respectively. As an extreme example, a tightly-packed horizontal row of circular bodies would exhibit high configurational anisotropy but no shape anisotropy. Meanwhile, an array of highly eccentric ellipses, all oriented horizontally but positioned randomly, would exhibit high shape anisotropy and low configurational anisotropy. We note that channelization is likely associated with {\em both} types of anisotropy. Certainly the overall configuration must support a channel, and, secondly, the geometry of the individual bodies that outline the channel could control its structure to some degree.

@@

As a second way to quantify medium anisotropy, we measure the tortuosity, $T_1$ and $T_2$, in the longitudinal and transverse directions respectively. Figure~\ref{fig4}(a), shows how both vary over the course of the simulation. As the bodies erode, $T_2$ grows significantly larger than $T_1$, indicating that passive tracers must
follow more tortuous paths when traveling transversely compared to longitudinally. The ratio $(T_2-1)/(T_1-1)$ provides a second measure of anisotropy to complement the permeability-dependent quantity. We have chosen to subtract one in parenthesis so that the numerator and denominator both vanish when no bodies are present, as analogous to resistivity. As seen in Figure~\ref{fig4}(b), the ratio $(T_2-1)/(T_1-1)$ grows significantly during erosion, up to a peak of nearly 8. This ratio can be decomposed into configurational and shape components, exactly as is done for the anisotropy (i.e.~by replacing a collection of eroded bodies with circles of the same areas and centers of mass). Figure~\ref{fig4}(b) shows that the shape component is greater for the majority of the simulation, but near the end of the simulation, the configurational component is competitive.

@@

The previous section shows intriguing features to arise from the erosion of a single, random initialization of solid bodies. The observations naturally raise the question: are the trends specific to the particular simulation shown or do these features emerge robustly across different initial conditions? To answer this question, we now analyze the same quantities---permeability, tortuosity, and anisotropy---over an ensemble of simulations having different numbers of bodies and different random initializations. We analyze runs having $M$ = 20, 40, 60, 80, and 100 bodies, with at least three different instances of each. This data set represents roughly 2,100 hours of computational time.
